Are Delayed Issues Harder to Resolve? Revisiting Cost-to-Fix of Defects throughout the Lifecycle

### 2016-09-15 phaseDelay4Emse.tex
Updates from ShareLaTeX

This paper documents the above study and explores reasons for this  mismatch between this common rule of thumb and empirical data.
In  summary, DIE is not some constant across all projects. Rather, DIE might
be an historical relic that  occurs intermittently 
only in  certain kinds of projects.  This is a significant result since it predicts that  new development
processes that promise to faster retire more issues will have only a  limited return on investment, and that a long-held truth in software engineering can be defeated on a large scale. 
\end{abstract}

@@

The DIE theory is worth examination since it has been used as the basis for decision-making in software engineering. For example,  Basili and Boehm comment that, since the 1980s,  this effect
 \begin{quote}
``...has been a major driver in focusing
industrial software practice on thorough
requirements analysis and design,
on early verification and validation, and
on up-front prototyping and simulation
to avoid costly downstream fixes''~\cite{boehm01}.
\end{quote}
 \begin{figure}[!b]  
\begin{center}
\includegraphics[,width=4in]{img/b81.png}
 \end{center}
 \caption{A widely-recreated chart of the DIE effect. Adapted from Boehm'81~\cite{Boehm81}. }\label{fig:b81}
 \end{figure}

Like any good theory, DIE includes a rationale for why the expected results would be seen. McConnell mentions it as a ``common observation'' in the field and  summarizes the intuitive argument for why it should be so: 
\begin{quote}
''A small mistake in upstream work can affect large amounts of downstream work. A change to a single sentence in a requirements specification can imply changes in hundreds of lines of code spread across numerous classes or modules, dozens of test cases, and numerous pages of end-user documentation''~\cite{mcconnell01}. 
\end{quote}
Glass also endorses this rationale, asserting that ``requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' is ``really just common sense''~\cite{glass02}.  Other researchers
are just as adamant in asserting that that the delayed issue effect is a generally useful law of software engineering.
For example, what we call the delayed issued effect was listed at \#1 by Boehm and Basili in their ``Top 10 list'' of ``objective and quantitative data, relationships,
and predictive models that help
software developers avoid predictable pitfalls
and improve their ability to predict
and control efficient software projects''~\cite{boehm01}.

@@

The dilemma of failing truths in the face of new evidence is not particular to software engineering. 
% To be fair, 
% SE is  not the only
% field where practitioners hang on to persistent beliefs, even if the evidence
% for those beliefs is not strong.
The medical profession applies  many practices based on studies
that have been disproved. For example,
a  recent article
in the Mayo Clinic Proceedings~\cite{prasad13} found  
146 medical practices based on studies 
in year $i$, but which were  reversed by subsequent trials within years $i+10$.
Even when the evidence for or against a treatment or intervention is clear, medical providers and patients may not accept it~\cite{aschwanden10}.
Aschwanden warns that ``cognitive biases''  such as  confirmation bias (the tendency to look for evidence that supports what you already know and to ignore the rest)  influence how we process information~\cite{aschwanden15}.

The cognitive issues that complicate medicine are also found in software engineering.
Passos et al.~\cite{passos11} warn that developers
usually develop their own theories of what works and what doesn't work in creating software, based on experiences from a few past
projects. Too often, these theories are assumed to be general truisms with widespread applicability to future projects. They comment ``past experiences were taken into account without 
much consideration for their context''~\cite{passos11}.
The results of J{\o}rgensen \& Gruschke~\cite{jorgensen09} support Passos et al. In an empirical study of expert effort estimation, they report that the experts rarely use lessons
  from past projects to improve their future reasoning in effort estimation~\cite{jorgensen09}. 
 They note that,
when the experts
  fail to revise their beliefs, this leads to poor
 conclusions and software projects  (see examples in~\cite{jorgensen09}).
 A similar effect is reported by
Devanbu et al.~\cite{prem16}  who examined responses from 564 Microsoft software developers from around
the world, they found that  ``(a)~programmers do indeed have very
strong beliefs on certain topics; (b)~their beliefs are primarily formed
based on personal experience, rather than on findings in empirical
research; (c)~beliefs can vary with each project, but do not necessarily
correspond with actual evidence in that project''.
Devanbu et al. further  comment that ``programmers give personal experience
as the strongest influence in forming their opinions''. This is a troubling
result, especially given the above comments from Passos and  J{\o}rgensen et al.~\cite{passos11,jorgensen09} about how quickly practitioners form, freeze, and rarely revisit those opinions.

@@

\section{Motivation: ``DIE'' is commonly held, yet poorly documented}
\label{sect:belief}
% To assess the prevalence of DIE,   we conducted a survey of software engineers. If our surveyed practitioners make management decisions based on their
% understanding of SE theory, then the DIE  may well influence their decisions.
One reason that industrial practitioners and academics believe so strongly in the delayed issue effect is that it is often referenced
in the SE literature. 
%For example,
%we know that \fig{b81} has been presented to the graduate SE class at North Carolina State University, without quarrel or %critical comment, every year for the last decade.
Yet when we look at the literature, the evidence for
delayed issue effect is both very sparse and very old.

We examined the literature on the delayed issue effect through a combination of snowball sampling~\cite{wohlin2014guidelines} and database search. We searched Google Scholar for terms such as ``cost to fix'' and ``defect cost'' and ``software quality cost''. The majority of the search results discuss quality measurements, quality improvement, or the cost savings of phase-specific quality improvement efforts (e.g., heuristic test case selection vs. smoke testing). A systematic literature review of software quality cost research can be found in \cite{karg2011systematic}. Relatively few articles discuss cost-to-fix as a function of when the defect was injected or found. We also conducted a general Google search for the above terms. We found a number of website articles and blog postings on this topic, e.g., \cite{IfSQ2013,Soni2016,Parker2013,Gordon2016}. From these, we gathered additional citations for the delayed issue effect, the vast majority of which were secondary sources, e.g.,~\cite{Leffingwell96,mead2004software,mcconnell1996software,mcconnell01,Tassey2002,boehm2012}. Our literature search is not exhaustive, but our results yielded an obvious trend: nearly every citation to the delayed issue effect could be traced to the seminal \textit{Software Engineering Economics}~\cite{Boehm81} or its related works~\cite{boehm88,boehm01}.\footnote{For example, popular sources such as~\cite{pressman2005software, boehm01,glass02,endres03} have a combined citation count of over 14,500 on Google Scholar can all trace their evidence to \textit{Software Engineering Economics}~\cite{Boehm81}.}

@@

Some studies that report less-than large
 increase in the effort required to fix delayed issues.  Boehm~\cite{Boehm80} provides data suggesting that the cost-to-fix curve for small projects
 %, two student projects of 2000 deliverable source instructions, 
 is flatter than for large projects (the dashed line of Figure~\ref{fig:cost-to-fix}). Data from NASA's Johnson Space Flight Center, reported by Shull~\cite{Shull02}, found that the cost to fix certain non-critical classes of defects was fairly constant across lifecycle phases (1.2 hours on average early in the project, versus 1.5 hours late in the project). Royce~\cite{Royce98} studied  a million-line, safety-critical missile defense system. Design changes (including architecture changes) required approximately twice the effort of implementation and test changes, and the cost-to-fix in implementation and test phases increased slowly. Boehm~\cite{Boehm10} attributes this success to a development process focused on removing architecture risk early in the lifecycle. Willis et al. (\cite{willis1998hughes}, page 54) provide tables summarizing the effort to fix over 66,000 defects as a function of lifecycle phase injected and removed from multiple projects. The tables are partly obscured, but seem to provide the first large scale evidence that a)~DIE need not be exponential and b)~DIE need not be monotonically increasing.  Again, the data points from these studies are not available, and thus newer evidence both in favor of and contrary to the DIE cannot be evaluated.

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=2.5in]{img/defect_type_frequency.png}
\end{center}
\caption{Relative frequencies of these defect types seen in our TSP data.}\label{fig:dtypes}.
\end{figure}
In our TSP data,  the relative frequencies of these defect types are shown in \fig{dtypes}. Around a quarter of the fixes were simple documentation changes.
That said, 75\% of the changes are quote  elaborate;   e.g. fixes to function necessitates a careful reflection of the purpose of the code.

@@

One or more defects are reported against a single plan item in the time tracking logs, e.g., a review session, an inspection meeting, a test execution. The time-tracking logs against the plan items for a defect includes the time required to (a)~collect data and realize there is an error, 
(b)~prepare a fix,  and (c)~apply some validation
procedure to check the fix (e.g. discuss it with a colleague or execute some tests).
Since multiple defects can be recorded against a plan item, the time-to-fix a defect is defined as:
\[
\text{\emph{time-to-fix a defect}} := \frac{\text{time for defect plan item}}{\text{\# of defects in plan item}}
\]

\subsubsection{Definition: development phase} \label{development_phase}
\begin{figure}[!b]  
\begin{center}
\includegraphics[width=1.6in]{img/wfall.png}  
\end{center}
\caption{Phases of our data.
Abbreviations: 
{\em Before}= before development; 
{\em Reqts}	  = requirements; 
{\em HLD}	  = high-level design; 
{\em IntTest} = Integration testing (with code from others); 
{\em SysTest} = system test (e.g. load stress tests); 
{\em AcceptTest}  = acceptance testing (with users); 
{\em review}        = private activity; 
{\em inspect}        = group activity.}
\label{fig:waterfall}
\end{figure}

@@

 The specific process work flows and practices are developed by the development team personnel who have received specific training on defining work processes as part of their Personal Software Process training. The process data was collected by the team members to self-manage their personal and team work. Them members also exhibited self-management behavior by estimating planning and scheduling the work tasks.

While the processes and work flows among these projects can vary, the logical order is described section \ref{development_phase} is followed. Development tasks such as requirements development, design, or code, are typically followed by an appraisal phase such as personal review or inspection. Effort and effectiveness of these activities vary among projects and developers.

@@

 The project schedule, cost, and scope  are characterized by calendar duration, development team size project, and product size (measured in added and modified lines of code and number of components).  These data are all available from the project fact sheets for each project. Summary statistics and the year of project initiation are displayed in Figure~\ref{fig:Project-descriptive-stats}. From this table we can make some observations about the range of project characteristics .

 \begin{figure}[ht]
\scriptsize
\centering
\begin{tabular}{lrrrrrrrl}
                                                                   & N   & Min & Q1   & Median & Q3    & Max   & Mean   &  Distribution \\\hline
Team size                                                          & 171 & 1   & 4    & 6      & 10    & 36    &  7.8    &  \includegraphics[width=15mm]{img/team_size.png} \\ 
Duration {[}days{]}                                                & 171 & 7   & 33   & 61     & 118   & 1918  & 107    & \includegraphics[width=15mm]{img/duration.png} \\
Added \& Modified LOC                                                 & 117 & 2   & 1125 & 4201   & 13092 & 88394 & 10259  & \includegraphics[width=15mm]{img/LOC.png} \\
Components                                                         & 171 & 8   & 26   & 49     & 107   & 4170  & 116.2  & \includegraphics[width=15mm]{img/components.png} \\
Year                                                               & 171 &     & 2011 & 2012   & 2013  & 2014  & 2011.9 & \includegraphics[width=15mm]{img/year.png}
\end{tabular}

@@

 Of the 171 projects in the sample, only 117 collected size data in lines of code, however all projects tracked effort and  the component counts with applied effort are provided. Other data is complete for all 171 projects. The projects were mostly of short duration and small to medium size. The median project began in 2012 lasted 61 days, produced 4200 Lines of Code, 49 components (modules or features). Duration ranged from 7 to 1918 days. Size ranged from minimal (this may represent a short maintenance project) to 88394. The earliest project was in 2006 and the most recent in 2014.

How many of these teams could be classified as "agile" is not clear because actual practices in the agile world can vary. We did not ask teams to self-identify, however we offer the following observations regarding characteristics commonly associated with agile behavior. 
\begin{itemize}
    \item all teams were self managed, defining work flows, practices, and schedules
    \item teams meet at least weekly to evaluate progress and replan
    \item most teams were small with a median size of 6 and a mean of 7.8. only 25\% of the teams were larger than 10 with a long tail on the distribution
    \item the median project lasted only 60 days, suggesting limited scope for each integration
\end{itemize}

\begin{figure}[!t] 
\vspace{0.5cm}
\begin{center}
\includegraphics[height=2.5in]{img/fix-phase-dist.pdf}
\end{center} 
\caption{Distribution of defects found and fixed by phase.}
\label{fig:fix-phase-dist}
\end{figure}

@@

    In the following presentation of our results, three statistical methods were used to test for the delayed
    issue effect: the Scott-Knott ranker;   bootstrap sampling (to test for statistical significantly
    different results); and an effect size test (to reject any significant differences that are trivially small).

    The Scott-Knott ranker 
      was   recommended  by  Mittas and Angelis in a
    recent TSE'13 article~\cite{mittas13} and by Ghotra et al. in a recent
    ICSE'15 article~\cite{ghotra2015icse}
    Scott-Knott is a 
    top-down clustering approach used to rank different treatments. If that 
    clustering finds an ``interesting division'' of the data, then some 
    statistical test is applied to the two divisions to check if they are 
    statistically significant different. If so, Scott-Knott considers recurses into both 
    halves.
    Before Scott-Knot recurses,  however,  it apples some  statistical hypothesis test $H$ to check
    if $m,n$ are significantly different.

    To operationalize ``interesting'', Scott-Knott seeks the division of 
    of $l$ treatments into subsets of size $m,n$ of sizes $ls,ms,ns$ and median values
    $l.\mu, m.\mu, n.\mu$ (respectively) in order to maximize
 $\frac{ms}{ls}abs(m.\mu - l.\mu)^2 + \frac{ns}{ls}abs(n.\mu - l.\mu)^2$.

   To operationalize $H$, we use both  bootstrap sampling  and the A12 effect size test.
   In other 
    words, we divide the data if \textit{both} bootstrap sampling and effect 
    size test agree that a division is statistically significant (with a 
    confidence of 99\%) and not a small effect ($A12 \ge 0.6$).
    For a justification of the use of non-parametric bootstrapping, see Efron 
    \& Tibshirani~\cite[p220-223]{efron93}. For a justification of the use of 
    effect size tests see Shepperd and MacDonell~\cite{shepperd12a}; 
    Kampenes~\cite{kampenes07}; and Kocaguenli et 
    al.~\cite{Kocaguneli2013:ep}. These researchers warn that even if a 
    hypothesis test declares two populations to be ``significantly'' 
    different, then that result is misleading if the ``effect size'' is very 
    small. Hence, to assess the performance differences we first must rule out 
    small effects using A12, a test   recently endorsed by Arcuri and 
    Briand at ICSE'11~\cite{arcuri11}.

    To  apply Scott-Knott, we divided data into the phases $P_0$ where issues are introduced. Next, for each division, we separated all the issues that were removed at different subsequent issues $P_r \in \{P_1,P_2,..\}$.
    For each pair $P_0,P_r$, we build one treatment containing the  issue resolution times for     issues raises in $P_0$ and resolved in $P_r$. These treatments
    were then ranked by Scott-Knott.

>>>>

This paper documents the above study and explores reasons for this  mismatch between this common rule of thumb and empirical data.
In  summary, DIE is not some constant across all projects. Rather, DIE might
be an historical relic that  occurs intermittently 
only in  certain kinds of projects.  This is a significant result since it predicts that  new development
processes that promise to faster retire more issues will not have a guaranteed return on investment (depending on the context where applied), and that a long-held truth in software engineering should not be considered a global truism. 
\end{abstract}

@@

The DIE theory is worth examination since it has been used as the basis for decision-making in software engineering. For example,  Basili and Boehm comment that, since the 1980s,  this effect
 \begin{quote}
``...has been a major driver in focusing
industrial software practice on thorough
requirements analysis and design,
on early verification and validation, and
on up-front prototyping and simulation
to avoid costly downstream fixes''~\cite{boehm01}.
\end{quote}
 \begin{figure}[!b]  
\begin{center}
\includegraphics[,width=4in]{b81.png}
 \end{center}
 \caption{A widely-recreated chart of the DIE effect. Adapted from Boehm'81~\cite{Boehm81}. }\label{fig:b81}
 \end{figure}

Like any good theory, DIE includes a rationale for why the expected results would be seen. McConnell mentions it as a ``common observation'' in the field and  summarizes the intuitive argument for why it should be so: 
\begin{quote}
''A small mistake in upstream work can affect large amounts of downstream work. A change to a single sentence in a requirements specification can imply changes in hundreds of lines of code spread across numerous classes or modules, dozens of test cases, and numerous pages of end-user documentation''~\cite{mcconnell01}. 
\end{quote}
Glass also endorses this rationale, asserting that ``requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' is ``really just common sense''~\cite{glass02}.  Other researchers
are just as adamant in asserting that the delayed issue effect is a generally useful law of software engineering.
For example, what we call the delayed issued effect was listed at \#1 by Boehm and Basili in their ``Top 10 list'' of ``objective and quantitative data, relationships,
and predictive models that help
software developers avoid predictable pitfalls
and improve their ability to predict
and control efficient software projects''~\cite{boehm01}.

@@

The dilemma of updating truths in the face of new evidence is not particular to software engineering. 
% To be fair, 
% SE is  not the only
% field where practitioners hang on to persistent beliefs, even if the evidence
% for those beliefs is not strong.
The medical profession applies  many practices based on studies
that have been disproved. For example,
a  recent article
in the Mayo Clinic Proceedings~\cite{prasad13} found  
146 medical practices based on studies 
in year $i$, but which were  reversed by subsequent trials within years $i+10$.
Even when the evidence for or against a treatment or intervention is clear, medical providers and patients may not accept it~\cite{aschwanden10}.
Aschwanden warns that ``cognitive biases''  such as  confirmation bias (the tendency to look for evidence that supports what you already know and to ignore the rest)  influence how we process information~\cite{aschwanden15}.

The cognitive issues that complicate medicine are also found in software engineering.
Passos et al.~\cite{passos11} warn that developers
usually develop their own theories of what works and what doesn't work in creating software, based on experiences from a few past
projects. Too often, these theories are assumed to be general truisms with widespread applicability to future projects. They comment ``past experiences were taken into account without 
much consideration for their context''~\cite{passos11}.
The results of J{\o}rgensen \& Gruschke~\cite{jorgensen09} support Passos et al. In an empirical study of expert effort estimation, they report that the experts rarely use lessons
  from past projects to improve their future reasoning in effort estimation~\cite{jorgensen09}. 
 They note that,
when the experts
  fail to revise their beliefs, this leads to poor
 conclusions and software projects  (see examples in~\cite{jorgensen09}).
 A similar effect is reported by
Devanbu et al.~\cite{prem16}  who examined responses from 564 Microsoft software developers from around
the world; they found that  ``(a)~programmers do indeed have very
strong beliefs on certain topics; (b)~their beliefs are primarily formed
based on personal experience, rather than on findings in empirical
research; (c)~beliefs can vary with each project, but do not necessarily
correspond with actual evidence in that project.''
Devanbu et al. further  comment that ``programmers give personal experience
as the strongest influence in forming their opinions.'' This is a troubling
result, especially given the above comments from Passos and  J{\o}rgensen et al.~\cite{passos11,jorgensen09} about how quickly practitioners form, freeze, and rarely revisit those opinions.

@@

\section{Motivation: ``DIE'' is commonly held, yet poorly documented}
\label{sect:belief}
% To assess the prevalence of DIE,   we conducted a survey of software engineers. If our surveyed practitioners make management decisions based on their
% understanding of SE theory, then the DIE  may well influence their decisions.
One reason that industrial practitioners and academics believe so strongly in the delayed issue effect is that it is often referenced
in the SE literature. 
%For example,
%we know that \fig{b81} has been presented to the graduate SE class at North Carolina State University, without quarrel or %critical comment, every year for the last decade.
Yet when we look at the literature, the evidence for
delayed issue effect is both very sparse and very old.
As  shown in this section,
a goal of agile methods 
is to reduce the difficulty associated with making changes later in the lifecycle~\cite{beck00}. Yet, as shown below,
relatively little empirical data exists on this point.

We examined the literature on the delayed issue effect through a combination of snowball sampling~\cite{wohlin2014guidelines} and database search. We searched Google Scholar for terms such as ``cost to fix'' and ``defect cost'' and ``software quality cost''. The majority of the search results discuss quality measurements, quality improvement, or the cost savings of phase-specific quality improvement efforts (e.g., heuristic test case selection vs. smoke testing). A systematic literature review of software quality cost research can be found in \cite{karg2011systematic}. Relatively few articles discuss cost-to-fix as a function of when the defect was injected or found. We also conducted a general Google search for the above terms. We found a number of website articles and blog postings on this topic, e.g., \cite{IfSQ2013,Soni2016,Parker2013,Gordon2016}. From these, we gathered additional citations for the delayed issue effect, the vast majority of which were secondary sources, e.g.,~\cite{Leffingwell96,mead2004software,mcconnell1996software,mcconnell01,Tassey2002,boehm2012}. Our literature search is not exhaustive, but our results yielded an obvious trend: nearly every citation to the delayed issue effect could be traced to the seminal \textit{Software Engineering Economics}~\cite{Boehm81} or its related works~\cite{boehm88,boehm01}.\footnote{For example, popular sources such as~\cite{pressman2005software, boehm01,glass02,endres03}, with a combined citation count of over 14,500 on Google Scholar, can all trace their evidence to \textit{Software Engineering Economics}~\cite{Boehm81}.}

@@

Some studies report smaller
 increases in the effort required to fix delayed issues.  Boehm~\cite{Boehm80} provides data suggesting that the cost-to-fix curve for small projects
 %, two student projects of 2000 deliverable source instructions, 
 is flatter than for large projects (the dashed line of Figure~\ref{fig:cost-to-fix}). Data from NASA's Johnson Space Flight Center, reported by Shull~\cite{Shull02}, found that the cost to fix certain non-critical classes of defects was fairly constant across lifecycle phases (1.2 hours on average early in the project, versus 1.5 hours late in the project). Royce~\cite{Royce98} studied  a million-line, safety-critical missile defense system. Design changes (including architecture changes) required approximately twice the effort of implementation and test changes, and the cost-to-fix in implementation and test phases increased slowly. Boehm~\cite{Boehm10} attributes this success to a development process focused on removing architecture risk early in the lifecycle. Willis et al. (\cite{willis1998hughes}, page 54) provide tables summarizing the effort to fix over 66,000 defects as a function of lifecycle phase injected and removed from multiple projects. The tables are partly obscured, but seem to provide the first large scale evidence that a)~DIE need not be exponential and b)~DIE need not be monotonically increasing.  Again, the data points from these studies are not available, and thus newer evidence both in favor of and contrary to the DIE cannot be evaluated.

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=2.5in]{defect_type_frequency.png}
\end{center}
\caption{Relative frequencies of these defect types seen in our TSP data.}\label{fig:dtypes}.
\end{figure}
In our TSP data,  the relative frequencies of these defect types are shown in \fig{dtypes}. Around a quarter of the fixes were simple documentation changes.
That said, 75\% of the changes are quite elaborate;   e.g. fixes to function necessitates a careful reflection of the purpose of the code.

@@

In the TSP, defect data includes the affected artifact, the estimated developer fix effort (find and fix), the lifecycle phases in which the defect was injected and removed, and the developer who implemented the fix. In the database, the task is associated with a plan item. Defects (one or more) are recorded in the defect log and associated with the plan item (task) in the time tracking logs. For example, a review session, an inspection meeting, or a test would be plan items associated with some product component. When defects are found and fixed, the time recorded in the time-tracking logs against the plan items includes the direct effort time (stop watch rather than wall clock time) required to (a)~collect data and realize there is an error, 
(b)~prepare a fix,  and (c)~apply some validation
procedure to check the fix (e.g. discuss it with a colleague or execute some tests). Although we have explicit estimates of "find and fix" effort for each defect, this fails to account for the full costs (e.g. meeting time or test execution). Because the vast majority of defects are removed in explicit removal phases, we chose to estimate defect cost using the entire time in removal phases divided by the number of defects. We recognize that this approach can exaggerate cost per defect for cases with few defects and large overhead effort, such large test suites or slow running tests that require continuous developer attention.  Nonetheless, this approach provides a better comparison between early removals from inspections later removals from test. The result will be a time per defect that is greater than the directly measured "find and fix" time, but smaller than the wall clock or calendar time.

Since multiple defects can be recorded against a plan item, the time-to-fix a defect is defined as:
\[
\text{\emph{time-to-fix a defect}} := \frac{\text{time for defect plan item}}{\text{\# of defects in plan item}}
\]

\subsubsection{Definition: development phase} \label{development_phase}
\begin{figure}[!b]  
\begin{center}
\includegraphics[width=1.6in]{wfall.png}  
\end{center}
\caption{Phases of our data.
Abbreviations: 
{\em Before}= before development; 
{\em Reqts}	  = requirements; 
{\em HLD}	  = high-level design; 
{\em IntTest} = Integration testing (with code from others); 
{\em SysTest} = system test (e.g. load stress tests); 
{\em AcceptTest}  = acceptance testing (with users); 
{\em review}        = private activity; 
{\em inspect}        = group activity.}
\label{fig:waterfall}
\end{figure}

@@

 The specific process work flows and practices are developed by the development team personnel who have received specific training on defining work processes as part of their Personal Software Process training. The process data was collected by the team members to self-manage their personal and team work. The members also exhibited self-management behavior by estimating planning and scheduling the work tasks.

While the processes and work flows among these projects can vary, the logical order described in section \ref{development_phase} is followed. Development tasks such as requirements development, design, or code, are typically followed by an appraisal phase such as personal review or inspection. Effort and effectiveness of these activities vary among projects and developers.

@@

 The project schedule, cost, and scope  are characterized by calendar duration, development team size project, and product size (measured in added and modified lines of code and number of components).  These data are all available from the project fact sheets for each project. Summary statistics and the year of project initiation are displayed in Figure~\ref{fig:Project-descriptive-stats}. From this table we can make some observations about the range of project characteristics.

 \begin{figure}[ht]
\scriptsize
\centering
\renewcommand{\arraystretch}{1}%
\begin{tabular}{lrrrrrrrl}
                                        & N   & Min & Q1   & Median & Q3    & Max   & Mean   &  Distribution \\\hline
Team size                               & 171 & 1   & 4    & 6      & 10    & 36    &  7.8    &  \includegraphics[width=25mm]{team_size.png} \\ 
Duration {[}days{]}                     & 171 & 7   & 33   & 61     & 118   & 1918  & 107    & \includegraphics[width=25mm]{duration.png} \\
Added \& Modified LOC                   & 117 & 2   & 1125 & 4201   & 13092 & 88394 & 10259  & \includegraphics[width=25mm]{LOC.png} \\
Defects Found \& Fixed                  & 171 & 1   & 28   & 95     & 278   & 4580  & 324.4  & \includegraphics[width=25mm]{defect_by_project.png} \\
Components                              & 171 & 8   & 26   & 49     & 107   & 4170  & 116.2  & \includegraphics[width=25mm]{components.png} \\
Project Initiation Year                 & 171 &    2006 & 2011 & 2012   & 2013  & 2014  & 2011.9 & \includegraphics[width=25mm]{year.png}
\end{tabular}

@@

 Of the 171 projects in the sample, only 117 collected size data in lines of code. However all projects tracked effort and  the component counts with applied effort are provided. Other data are complete for all 171 projects. The projects were mostly of short duration and small to medium size. The median project began in 2012 lasted 61 days, produced 4,200 Lines of Code, 49 components (modules or features). Duration ranged from 7 to 1,918 days. Size ranged from minimal (this may represent a short maintenance project) to 88,394. The earliest project was in 2006 and the most recent in 2014.

How many of these teams could be classified as "agile" is not clear because actual practices in the agile world can vary. We did not ask teams to self-identify, however we offer the following observations regarding characteristics commonly associated with agile behavior. 
\begin{itemize}
    \item all teams were self managed, defining work flows, practices, and schedules
    \item teams met at least weekly to evaluate progress and re-plan
    \item most teams were small with a median size of 6 and a mean of 7.8; only 25\% of the teams were larger than 10 with a long tail on the distribution
    \item the median project lasted only 60 days, suggesting limited scope for each integration
\end{itemize}

\begin{figure}[!t] 
\vspace{0.5cm}
\begin{center}
\includegraphics[height=2.5in]{fix-phase-dist.pdf}
\end{center} 
\caption{Distribution of defects by phase removed.}
\label{fig:fix-phase-dist}
\end{figure}

@@

    In the following presentation of our results, three statistical methods were used to test for the delayed
    issue effect: the Scott-Knott ranker;   bootstrap sampling (to test for statistical significantly
    different results); and an effect size test (to reject any significant differences that are trivially small).
    Scott-Knott allows for a direct answer to the following questions:
    \bi
    \item
    Given an issue
    raised at phase $i$ and resolved at phase $\forall j,k \in \{i,i+1,i+2,...\}$,...
    \item 
    ... Is it
    true that  the time to resolve issues in phase $j$ is significantly different 
    to the time to resolve issues in phase $k$? 
    \ei
    Note that if $j,k$ times are significantly different, then we can compare
    the median values to say (e.g.) resolution time at phase $k$ is 3 times slower than phase $j$.
    Note also that if all times $j,k$ are {\em not} significantly different
    then we say that the phases all {\em rank} the same (and we denote this by setting
    all such {\em ranks} to 1).

    In the following results, we nearly always encountered the second case;
    i.e.   the times to resolve issues at different times were usually {\em not significantly}
    different.

    As to technical details of the Scott-Knott methods, this ranker 
      was   recommended  by  Mittas and Angelis in a
    recent TSE'13 article~\cite{mittas13} and by Ghotra et al. in a recent
    ICSE'15 article~\cite{ghotra2015icse}
    Scott-Knott is a 
    top-down clustering approach used to rank different treatments. If that 
    clustering finds an ``interesting division'' of the data, then some 
    statistical test is applied to the two divisions to check if they are 
    statistically significant different. If so, Scott-Knott considers recurses into both 
    halves.
    Before Scott-Knot recurses,  however,  it applies some  statistical hypothesis test $H$ to check
    if $m,n$ are significantly different. 
    To operationalize ``interesting'', 
    \bi
    \item
    Scott-Knott seeks the division of 
    $l$ treatments into subsets of size $m,n$ (so if $n$ was 
    appended to the  end of $m$ then that new list
    would the same as $l$).
    \item
    We say that $l,m,n$ have
    sizes $ls,ms,ns$ and median values
    $l.\mu, m.\mu, n.\mu$ (respectively) 
    \item
    Scott-Knott tries all ways to split $l$ into $m,n$ and returns
    the one that maximizes the differences in the mean values
    before and after the splits; i.e. 
 \[\frac{ms}{ls}abs(m.\mu - l.\mu)^2 + \frac{ns}{ls}abs(n.\mu - l.\mu)^2\]
   \ei
   To operationalize $H$, we use both  bootstrap sampling  and Vargha and Delaney's A12 effect size test.
   In other 
    words, we divide the data if \textit{both} bootstrap sampling and effect 
    size test agree that a division is statistically significant (with a 
    confidence of 99\%) and not a small effect ($A12 \ge 0.6$).
    For a justification of the use of non-parametric bootstrapping, see Efron 
    \& Tibshirani~\cite[p220-223]{efron93}. For a justification of the use of 
    effect size tests see Shepperd and MacDonell~\cite{shepperd12a}; 
    Kampenes~\cite{kampenes07}; and Kocaguenli et 
    al.~\cite{Kocaguneli2013:ep}. These researchers warn that even if a 
    hypothesis test declares two populations to be ``significantly'' 
    different, then that result is misleading if the ``effect size'' is very 
    small. Hence, to assess the performance differences we first must rule out 
    small effects using Vargha and Delaney's A12 test, a test    endorsed by Arcuri and 
    Briand at ICSE'11~\cite{arcuri11}.

    To  apply Scott-Knott, we divided data into the phases $P_0$ where issues are introduced. Next, for each division, we separated all the issues that were removed at different subsequent issues $P_r \in \{P_1,P_2,..\}$.
    For each pair $P_0,P_r$, we build one treatment containing the  issue resolution times for     issues raised in $P_0$ and resolved in $P_r$. These treatments
    were then ranked by Scott-Knott.

\begin{figure} 
% \renewcommand{\baselinestretch}{0.8}
% \small
\begin{center}

%\resizebox{\columnwidth}{!}{
\scriptsize
\begin{tabular}{c|lr|rr|rl}
& \multicolumn{2}{c|}{ } & \multicolumn{2}{c|}{Percentiles} & \multicolumn{2}{c}{~ } \\ 
  & \multicolumn{2}{c|}{ } & \multicolumn{2}{c|}{(units = } & \multicolumn{2}{c}{Growth with respect to earliest phase  } \\ 
   & \multicolumn{2}{c|}{Phase} & \multicolumn{2}{c|}{  minutes)} & \multicolumn{2}{c}{(unitless  ratios of two time values) } \\\hline

  rank & injected & removed & 50th &   IQR & \multicolumn{2}{c}{50th percentile growth}  \\ 
\hline
